---
title: "Data Annotation"
author: "Alison Paquette"
date: "11/14/2016"
output: html_document
---

#Part 0A: Load All packages & Functions used in this analysis
```{r}
library(oligo) #For Affymetrix Data
library(lumi) #For Illumina Data
library(biomaRt) #To convert probes to genes
ensembl = useMart("ensembl",dataset="hsapiens_gene_ensembl") #Load Human Ensembl Library

outersect <- function(x, y) {
  sort(c(setdiff(x, y),
         setdiff(y, x)))
}

RemoveDupProbes<-function(IDs, NormData){
Report<-as.data.frame(matrix(NA,nrow = 4,ncol=1))
rownames(Report)<-c("Initial Probes","Probes w/ no Genes","Duplicated Probes","Unique Probes")

probeids=row.names(NormData)
Report[1,1]<-length(probeids)

ArrayProbes<-unique(IDs[,1])
Report[2,1]<-(length(probeids)-length(ArrayProbes))
  
  #Isolate Duplicated Probes
Duplicated<-subset(IDs,duplicated(IDs[,1])|duplicated(IDs[,1],fromLast=T)==T)
  
DupTable<-as.data.frame(table(Duplicated[,1]))
rownames(DupTable)<-as.character(DupTable$Var1)
  
  #Create list of Data that is Duplicated and subset this Data
Duplications<-rownames(DupTable)
Report[3,1]<-length(Duplications)
  
#Isolate Unique Probes
UniqueProbes<-outersect(ArrayProbes,Duplications)
Unique<-subset(IDs,duplicated(IDs[,1])==F)
rownames(Unique)<-as.character(Unique[,1])
  
if (dim(Unique)[1]!=length(ArrayProbes)){
  stop("Number of Unique Probes Not Equal To Array Probes")
}
 
Unique<-(Unique[UniqueProbes,])
rownames(Unique)<-Unique[,1]
Report[4,1]<-length(rownames(Unique))

print(Report)
  #Create Dataset with these unique probes
Unique.Data<-merge(Unique,(NormData),by='row.names',all=F)
Unique.Data<-Unique.Data[,-1]
if (dim(Unique)[1]+(dim(DupTable)[1])!=length(ArrayProbes)){
  stop("Unique and Non Unique Probes do NOT equal Probe IDs")
} 
(Unique.Data)
}

CondenseProbes<-function(ProbeData1){

rownames(ProbeData1)<-ProbeData1[,1]
Ensembl_IDs<-list(as.character(ProbeData1[,2]))
TempData<-ProbeData1[,-c(1:3)]
CollapsedData<-aggregate(TempData,by=Ensembl_IDs,FUN=mean,na.rm=TRUE, na.action=NULL)
rownames(CollapsedData)<-CollapsedData[,1]
CollapsedData<-CollapsedData[,-1]
  
}

```


#Part OB: Load covariate data and identify what samples to use
```{r}
Covar<-read.csv("~/Documents/Project 1. Preterm Birth/PaperPreprocessingJan/Preprocessing/1262016UpdatedCovar.csv")

rownames(Covar)<-as.character(Covar$SampleIdentifier)

#We have elected to exclude indivdiuals with Preeclampsia
NoPE<-subset(Covar,!PETstatus=="Y"|is.na(PETstatus))
NoCA<-subset(NoPE,!CA.Diagnosis=="Yes"|is.na(CA.Diagnosis))
NoInbetween<-subset(NoCA,!is.na(birth.type))

Covar2<-subset(NoInbetween,!StudyIdentifier=="GSE44711") #Need to dop this one study because it only  has preterm infants
Covar2$StudyIdentifier<-as.character(droplevels(Covar2$StudyIdentifier))
table(Covar2$StudyIdentifier)

Preterm<-subset(Covar2,birth.type=="PRETERM")
Term<-subset(Covar2,birth.type=="TERM")
rbind(table(Preterm$StudyIdentifier),table(Term$StudyIdentifier))


```

**Example 1: : HuGene-2_0-st (Gene Version)**
#Part 1: Loading .CEL Files of interest
```{r}
#Subset Files you want from Covariate
X<-"GSE73374"
SampleID<-as.character(rownames(subset(Covar2,StudyIdentifier==X)))

#Data munging: Getting .CEL messy filenames to The actual filenames
setwd("~/Documents/Project 1. Preterm Birth/GSE73374 Raw Data/GSE73374_RAW/")
filenames <- list.celfiles()
filenames.tmp<-as.data.frame(strsplit(filenames, split="_"))
filenames.tmp2<-cbind(t(filenames.tmp),filenames)
rownames(filenames.tmp2)<-filenames.tmp2[,1]
filenames.tmp2<-filenames.tmp2[SampleID,]
filenames<-filenames.tmp2[,3]

rawData <- read.celfiles(filenames)
```

#Part 2: Preprocess/Prepare Datasets
```{r}
NormData <- rma(rawData)#Note: The output of RMA using Oligo is already log2 normalized (unlike xps, previous packages, as described in https://www.bioconductor.org/packages/release/bioc/vignettes/oligo/inst/doc/oug.pdf)

par(mfrow=c(2,1))
boxplot(exprs(rawData),main=c(X,"Raw"))
boxplot(exprs(NormData),main=c(X,"Normalized"))
NormData<-exprs(NormData)
```

#Part 3: Convert Probes to Genes
```{r}
IDs<-getBM(filters="affy_hugene_2_0_st_v1",attributes= c("affy_hugene_2_0_st_v1","ensembl_gene_id","hgnc_symbol"),values=row.names(NormData), mart=ensembl)

ProbeData1<-RemoveDupProbes(IDs,NormData)

GSE73374<-CondenseProbes(ProbeData1)

#save(GSE73374,file="~/Documents/Project 1. Preterm Birth/PaperPreprocessingJan/Preprocessing/GSE733374_12282016.RData")
```

**GSE75010: HuGene-1_0-st (Gene Version)**
#Part 1: Load raw data files (downloaded from GEO here: )
This is analogous to Issue 2: Extract Data from studies (Ramasamy et al, 2008)
```{r}
#Subset Files you want from Covariate
X<-"GSE75010"
SampleID<-as.character(rownames(subset(Covar2,StudyIdentifier==X)))

#Data munging: Getting .CEL messy filenames to The actual filenames
setwd("~/Documents/Project 1. Preterm Birth/GSE75010 Raw Data/")
filenames <- list.celfiles()
filenames.tmp<-as.data.frame(strsplit(filenames, split="_"))
filenames.tmp2<-cbind(t(filenames.tmp),filenames)
rownames(filenames.tmp2)<-filenames.tmp2[,1]
filenames.tmp2<-filenames.tmp2[SampleID,]
filenames<-filenames.tmp2[,6]

rawData <- read.celfiles(filenames)

colnames(rawData)<-SampleID
```

#Part 2: Preprocess/Prepare Datasets
This is analgous to Issue 3: Prepare datasets from different platforms (Ramasamy et al, 2008)
```{r}
NormData <- rma(rawData)

#Quick QC of Data
par(mfrow=c(2,1))
boxplot(exprs(rawData),main=c(X,"Raw"))
boxplot(exprs(NormData),main=c(X,"Normalized"))

NormData<-exprs(NormData)
```

```{r}
IDs<-getBM(filters="affy_hugene_1_0_st_v1",attributes= c("affy_hugene_1_0_st_v1","ensembl_gene_id","hgnc_symbol"),values=rownames(NormData), mart=ensembl)

Probeset1<-RemoveDupProbes(IDs,exprs(NormData))
GSE75010<-CondenseProbes(Probeset1)

save(GSE75010,file="~/Documents/Project 1. Preterm Birth/PaperPreprocessingJan/Preprocessing/GSE75010_12282016.RData")
```

